---
title: "Lahman Database Model"
author: "Darko Bergant"
date: "2017-10-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About

This tutorial shows how to use
[datamodelr](https://github.com/bergant/datamodelr) to supplement R package
documentation with relational diagrams.

Examples are using data frames from
[Lahman](https://cran.r-project.org/package=Lahman) package.


## Extract meta data 

There are > 20 data frames in the package. We can put them in a list
`dataframes`:

```{r}
lahman <- new.env()
table_names <- data(package="Lahman")$results[,"Item"]
data(list = table_names, package = "Lahman", envir = lahman)
dataframes <- as.list(lahman)
str(dataframes, max.level = 1)
```

**datamodelr** can create data model object from a data frame with columns:
`table`, `column`, `key`, `ref` and `ref_col` ("meta data" structure). See
`?as.data_model` for more info.

We can create meta data directly from the Lahman data frames:

```{r metadata, message=FALSE}
library(dplyr)
library(purrr)

meta_data <- 
  dataframes %>% 
  map(~data_frame(
    column = names(.),
    type = map_chr(., ~paste(class(.), collapse = ",") ),
    key = NA,
    ref = NA,
    ref_col = NA
  )) %>% 
  bind_rows(.id = "table") 

glimpse(meta_data)

```

## Column descriptions

It is nice to have column descriptions in data model. Source: http://seanlahman.com/files/database/readme2016.txt, section 2.0


```{r}
column_descriptions <-
  read.table("data/column_descriptions.csv", 
             header = TRUE, stringsAsFactors = FALSE, sep = ",")

meta_data <-
  meta_data %>% 
  left_join(column_descriptions, by = c("table", "column")) %>% 
  mutate(description = substring(description, 1, 30))

```


Use `as.data_model` to convert meta data to data model object:

```{r}
library(datamodelr)
dm <- as.data_model(meta_data)
dm
```

Create diagram for selected tables:
 
```{r simple_graph, fig.height=9}
selected_tables = c("Master", "Batting", "Pitching", "Fielding", "Teams")
g <- dm_create_graph(dm, focus = list(tables = selected_tables), 
                     col_attr = c("column", "type"))
dm_render_graph(g)
```



## Add keys 

We could use `dm_set_key` and `dm_add_references` to add keys and references
directly to a data model object. But, it looks verbose when managing large number
of tables and references. Sometimes it is easier to manipulate the meta data
before coercing it to the data model object.


```{r add_keys}

# simple key rules structure;

key_rules = list(
  list(
    table = "Master",
    column = "playerID"
  ),
  list(
    table = c("Batting", "Fielding", "Pitching"),
    column = c("playerID", "yearID", "teamID", "lgID")
  ) , 
  list(
    table = "Teams",
    column = c("yearID", "teamID", "lgID")
  )  
)

# use dplyr::mutate to update the key column
for(key_rule in key_rules) {
  meta_data <- 
    mutate(meta_data, 
      key = ifelse(table %in% key_rule$table & 
        column %in% key_rule$column, TRUE, key))
}

dm <- as.data_model(meta_data)
```

Use "keys_only" view type to render diagram with only key columns:
```{r diagram_keys, fig.height=2}
g <- dm_create_graph(dm, focus = list(tables = selected_tables),
                     view_type = "keys_only", col_attr = c("column", "type"))

dm_render_graph(g)
```

# Add references

Mark relations with column `ref` (related table) and `ref_cols` referenced columns
if related table has multiple columns in a key.

In this example the relations to "Master" and "Teams" are created from
"Batting", "Fielding" and "Pitching":

```{r add_references}

ref_rules <- list(
  list(
    table = c("Batting", "Fielding", "Pitching"),
    column = "playerID",
    ref = "Master"
  ),
  list(
    table = c("Batting", "Fielding", "Pitching"),
    column = c("yearID", "teamID", "lgID"),
    ref = "Teams"
  )
)


# update meta data
for(ref_rule in ref_rules) {
  meta_data <- mutate( 
    meta_data, 
    ref = ifelse(table %in% ref_rule$table & column %in% ref_rule$column, 
                 ref_rule$ref, ref),
    ref_col = ref # same column names in referenced table
  )  
}

dm <- as.data_model(meta_data)
dm


```


```{r diagram_refs, fig.height=2}
g <- dm_create_graph(
  dm,
  focus = list(tables = selected_tables),
  view_type = "keys_only",
  col_attr = c("column", "type")
)

dm_render_graph(g)



```

With all columns selected, the diagram would be like:

 
```{r fig.height=10} 
g <- dm_create_graph(
  dm,
  focus = list(tables = selected_tables),
  view_type = "all",
  col_attr = c("column", "type", "description")
)

dm_render_graph(g)


```


# Table overview



**Batting** table:

```{r}
DT::datatable(
  dm$columns %>% 
    filter(table == "Batting") %>% 
    mutate(
      column = ifelse(key, sprintf("<strong>%s</strong>", column), column)
    ) %>% 
    select(column, description, type, ref), 
  filter = "none",
  escape = -1,
  rownames = FALSE,
  options = list(
    pageLength = 100,
    dom = 't'
  )
)
```

